\documentclass[11pt]{article}
\usepackage[margin=0.9in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{libertine}
\renewcommand{\sfdefault}{fvs} % Use as sans-serif
\renewcommand{\ttdefault}{fvm} % Use as monospace
\renewcommand{\ttfamily}{\small\fontfamily{\ttdefault}\selectfont} % Slightly smaller monospace font
\newcommand{\eg}{{\em e.g.}}
\newcommand{\coe}{CO$_2$e}
\newcommand{\gcoe}{g \coe}
\newcommand{\kgcoe}{k\gcoe}
\newcommand{\gcoekwh}{\gcoe\ / kWh}
%\usepackage{helvet} % Helvetica font for sans-serif
%\usepackage{courier} % Courier font for monospace
\usepackage{hyperref}
\hypersetup{
    hidelinks
}
\usepackage[style=numeric]{biblatex}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{titling}

\titleformat{\section}{\sffamily\Large}{\thesection}{1em}{}
\titleformat{\subsection}{\sffamily\large}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\sffamily\normalsize}{\thesubsubsection}{1em}{}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.5em}

\addbibresource{ref.bib}
\title{Modeling the impact of IT- and AI-related emissions in a university setting}
\author{Massimiliano Poletto\thanks{Independent consultant, max.poletto@gmail.com}, Nils Handler\thanks{University of Zürich, nils.handler@uzh.ch. Contact author.}}

\begin{document}

\maketitle

\section*{Abstract}
Information technology (IT) plays a major role in society and has non-negligible environmental impact, accounting for over 5\% of global electricity consumption and about 2\% of global carbon emissions. Moreover, artificial intelligence (AI) applications, especially large language models (LLMs), are being adopted at historically high rates. Their growth is expected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030. Therefore, an assessment of IT and AI emissions is a necessary component of any net-zero roadmap. As part of the University of Zürich's own net-zero efforts, we have attempted to quantify and analyze the impact of IT and AI on the university's carbon footprint. We present a simple interactive model of IT emissions on a university campus. The model approximates operational and lifecycle emissions of personal devices, university-owned computing and network infrastructure, cloud computing, and LLMs. The model is intended to provide university decision makers with emissions estimates under different scenarios, with sufficient accuracy to support effective decision-making. Our analysis shows that the embodied emissions of IT infrastructure and personal devices are a major component of overall IT emissions, that operational impacts are comparatively modest, and that LLM adoption, even under aggressive growth scenarios, is unlikely to comprise more than $5\%$ of overall IT-related emissions.

\section{Introduction}

Information technology accounts for over 5\% of global electricity consumption and at least 1.7\% of global carbon emissions~\cite{wb:itu:ict}. At the University of Zürich (UZH), which with over 38,000 students and staff is the largest Swiss university, IT in 2024 accounted for about 1.5 Mt \coe\ (megatons of carbon dioxide equivalent) of greenhouse gas (GHG) emissions, approximately 6.3\% of total university emissions in 2024~\cite{uzh:sustainability:report}. The cited report takes into account university-owned workstations, servers, and audiovisual equipment, but does include other IT-related activities such as networking, cloud computing, or the growing use of artificial intelligence, and in particular the adoption of generative AI and large language models (LLMs).

AI applications are being adopted at historically high rates, more quickly than the personal computer or the internet were adopted in the past~\cite{bick:ai:adoption}. Enterprise AI adoption has accelerated, with generative AI tools seeing widespread deployment across industries within months of their release~\cite{mckinsey2024ai}. In July 2025, ChatGPT received 2.5 billion queries per day, up from 1 billion in December 2024, and within an order of magnitude of Google's approximately 15 billion queries per day~\cite{techcrunch:chatgpt}. AI growth is currently projected to increase the share of global electricity consumption for data centers from $\sim 1\%$ in 2024 to $\sim 3\%$ by 2030~\cite{iea:ai:energy}.

The University of Zürich has set itself the target of becoming climate-neutral by 2030 within the categories of energy, transportation, food, and waste, with additional goals in 2035 and beyond~\cite{uzh:sustainability:report}. Given the current sizable contribution of IT to the university's emissions, and the potential for growth through AI adoption in the coming years, we felt it was necessary to understand IT emissions and their growth trajectories in more detail.

This paper presents a simple interactive model of IT emissions on a university campus. We developed it in the specific context of the University of Zürich, but believe that it is parameterizable and general enough to apply to other similar university campuses or large corporate facilities.

We seek to answer three broad questions:
\begin{itemize}
  \item What are the main contributors to the IT emissions of a university?
  \item Could migrating some on-campus IT services to commercial cloud providers decrease the overall campus footprint?
  \item What is the likely impact of the surge in AI adoption? In particular, will the emissions impact of AI be so large as to make other sources of IT emissions negligible?
\end{itemize}

Our model sets relatively broad system boundaries. It takes into account university-owned equipment but also university services hosted by commercial cloud providers and university use of commercial generative AI products. We have also parameterized it to account for student devices not owned by the university, since they are key to student work and constitute the bulk of the university's end-user devices. We attempt to model devices' full lifecycle impact, including manufacture and disposal. 

Experts rightly point to the many potential benefits of AI in the areas of climate change mitigation and adapation~\cite{climate:change:ai}. On a university campus, these might include more efficient building heating and cooling, improved transportation logistics, etc. However, those benefits are still to come and difficult to predict. In this paper, we consider only the {\em direct} impacts of AI in terms of increased electricity consumption and the resulting emissions.

The rest of this paper is organized as follows. Section~\ref{sec:model:methodology} presents the general modeling framework, including system boundaries and assumptions. Section~\ref{sec:model:details} justifies some of the specific parameter values used in the model. Section~\ref{sec:results} describes results and explores different scenarios and assumptions. Section~\ref{sec:recommendations} offers recommendations to university IT administrators. Section~\ref{sec:related:work} describes related work in this area.

\section{Modeling methodology}
\label{sec:model:methodology}

Our aim is to model energy use from university IT and AI broadly, without restricting ourselves to university-owned devices, specific IT departments, etc. At the same time, we want to be able to distinguish the contributions of different factors, such as student-owned devices or LLMs, and to compare the model's output against existing university data points (energy bills, etc.).

\subsection{System boundaries}

We define the boundaries of our model as follows.

\subsubsection*{Population}

We include IT-related activity from all members of the UZH community: faculty, staff, and students.

\subsubsection*{On-campus devices}

Student-owned devices, such as laptops and smartphones, are in scope, since they are the primary means by which students access university IT services and do their coursework. Given the duration of studies, we include personal devices' embodied emissions in the model.

University-owned devices are in scope. We have accurate numbers for staff workstations, classroom computers, network access points, and audiovisual devices managed by the central IT department. We can only estimate other devices, such as networking equipment or computers managed by individual research groups. As before, we include these devices' embodied emissions in the model.

\subsubsection*{Cloud and supercomputing facilities}

UZH at present does not make significant use of commercial cloud compute, but the model allows simulating the shift of some fraction of the server load to cloud VMs.

Due to lack of data, we do not attempt to model emissions from application-level cloud services such as Microsoft 365 or Google Workspace. This omission leads to undercounting total IT energy use but to overestimating the potential relative impact of AI.

UZH faculties do make use of shared off-campus supercomputing facilities, primarily the Swiss National Supercomputing Center, CSCS~\cite{cscs}. We estimate the operational impact of these workloads, but, again due to lack of data, we do not include embodied emissions. Again, this undercounts total university IT energy use but overestimates the potential relative impact of AI.

\subsubsection*{Large language models}

We estimate inference-time emissions of current large language models based on the literature. We also include training-time emissions and embodied emissions of language models (thouse these are harder to estimate and we err on the side of overestimating them).

\subsubsection*{Summary}

This table summarizes what is in scope in the current model. The main gap is emissions of cloud services such as Microsoft 365 or Goole Workspace.

\begin{center}
  \begin{tabular}{|l|c|c|}
  \hline
  \textbf{Device or service} & \textbf{Operation} & \textbf{Lifecycle} \\ \hline
  Student laptops \& phones & yes & yes \\
  University workstations, access points, etc. & yes & yes \\
  Cloud servers & yes & yes \\
  Cloud services & no & no \\
  Supercomputing facilities & yes & no \\
  LLMs & yes & yes \\ \hline
  \end{tabular}
  \label{tab:scopes}
  \end{center}

\subsection{Accounting approach}

\subsubsection*{On-campus devices}

We model campus device emissions bottom-up. We first identify categories of devices (phones, laptops, workstations, monitors, routers, etc.). For each device type, we research a range of common models, then use manufacturer data sheets or other available documentation to identify operational power draw (in watts) as well as embodied emissions related to manufacture and disposal (in \kgcoe) per device. Based on this data, we model the power draw and embodied emissions of an ``exemplar'' device of each type.

For each device type we also estimate (1) a count (\eg, a laptop for every student and staff), (2) an estimated duty cycle (\eg, laptops used 8 hours per day, 200 days per year), and (3) a useful lifetime (in years).

Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (\kgcoe).

We then estimate the local grid's emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number. We obtain emissions intensity data from Electricity Maps~\cite{electricitymaps}.

In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution. However, we have insufficient data on the precise form of such a distribution, so we limit our analysis to the linear mode.

\subsubsection*{Cloud and supercomputing facilities}

We attempt to model a hypothetical shift of some workloads from servers in an on-premise data center to a commercial cloud. The parameters under consideration include:

\begin{itemize}
  \item On-premise power usage effectiveness (PUE), the ratio of total data center power (including cooling) to power used for computation.
  \item Cloud data center PUE and grid emissions intensity (since the cloud data center may be located in a different region than the on-premise data center).
  \item An average conversion ratio of on-premise servers to cloud servers.
\end{itemize}

We measure the impact of cloud servers in terms of annual operational emissions and amortized emissions due to manufacturing and disposal, in units of \kgcoe\ / year.

As mentioned earlier, we do not attempt to model application-level cloud services.

For supercomputing facilities (in particular, CSCS) we rely on annual statistics of total power consumption for peak compute load and make assumptions about duty cycle. We do not have reliable data about the embodied emissions of the facility, but suspect that the fraction of these emissions that is attributable to UZH use is negligible relative to all other UZH emissions.

\subsubsection*{Large language models}

For LLMs, we estimate operational energy use (Wh per query) using current estimates in the literature of state-of-the-art models~\cite{devries2023growing,epoch2025howmuchenergydoeschatgptuse}.

We consider the training overhead and the embodied emissions of the servers and GPUs that the model runs as a fixed cost that is amortized across all queries to a model. As a result, in practice, this overhead is constantly shrinking. We incorporate it into the model as a configurable percentage overhead of the operational energy use.

\subsection{Parameters and assumptions}

The model's outputs are affected by several factors, and where we were unable to find good data we made assumptions that we considered reasonable. All of these parameters and assumptions are parameterizable by users of the model. We list these sources of uncertainty below.

\begin{description}
  \item[Cloud data center PUE] As of 2025, industry average power usage effectiveness is 1.56, with Google averaging 1.09 across its fleet~\cite{google:datacenter:efficiency} and Microsoft ranging from 1.12 to 1.30~\cite{microsoft:datacenter:efficiency}. We choose a PUE of 1.2 as representative of a state-of-the-art hyperscaler data center.
  \item[On-premise data center PUE] Based on limited UZH historical power data, we assume an on-premise PUE of 2.0, attributable to inefficient cooling.
  \item[Device duty cycle] We make unsupported but plausible assumptions about the number of days that different devices are used (\eg, end-user devices are not used every day) and their duty cycle throughut the day.
  \item[Electricity emissions intensity for on-premise data centers] For Switzerland, we can compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps}. Emissions vary from a high of 64 \gcoekwh\ in the middle of the night to a low of 45 \gcoekwh\ in the middle of the day. This variation is sufficiently large that we refine the model's device duty cycle to four six-hour periods over the course of the day. It allows us to specify, for example, that user-devices are mostly used during the day whereas servers run around the clock.
  \item[Electricity emissions intensity for cloud data centers and LLMs] These facilities are unlikely to be located in Switzerland, and Switzerland's grid is cleaner than average. We use the European Union's average emissions intensity of electricity generation in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
  \item[LLM usage] We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).
\end{description}

\section{Model details}
\label{sec:model:details}

The previous section describes our overall modeling approach, which we believe is broadly applicable to university and other large campus settings. This section presents the reasoning behind specific default values we employ in the UZH evaluation.

\subsection{On-campus devices}

\subsubsection*{Number of devices}

We have an authoritative census of devices (workstations, servers, audiovisual equipment) managed by central IT, as follows.

\begin{center}
  \begin{tabular}{|l|r|r|}
  \hline
  \textbf{Device type} & \textbf{Count} & \textbf{Life (y)}\\ \hline
  Laptops & 10,440 & 4 \\
  Monitors & 10,400 & 6 \\
  Servers & 1,121 & 5 \\
  Printers & 333 & 6 \\
  Mobile phones & 2,439 & 3 \\
  A/V devices & 431 & 6 \\ \hline
  \end{tabular}
  \label{tab:oncampusdevices}
\end{center}

In addition to these numbers, which cover staff and facilities, we assume that 80\% of the student population of 28,000 people have a laptop (or equivalent device, such as a tablet), and that everyone at the university has mobile phone, resulting in a rough estimate of 32,000 laptops and 35,000 mobile phones.

Traditional desktop computers appear to be infrequent on campus. By default we set their number to zero in our model, but we still provide parameters for them.

\subsubsection*{Operational emissions}

The Swiss electricity mix is substantially cleaner than the European or North American average, so generic manufacturer estimates of lifetime usage-related emissions are unsuitable. Unfortunately, in manufacturer literature we found surprisingly few sources of data on device power consumption, as opposed to estimated lifetime CO$_2$ emissions. Based on some manufacturer data sheets, primarily from Apple, Cisco, and Dell, we model power draw of different devices while under load using the following initial parameters:

\begin{center}
\begin{tabular}{|l|r|r|}
\hline
\textbf{Device} & \textbf{Mean (W)} & \textbf{$\sigma$ (W)} \\ \hline
Laptop & 30 & 5 \\ \hline
Monitor & 50 & 10 \\ \hline
Desktop & 100 & 20 \\ \hline
Server & 400 & 100 \\ \hline
Mobile phone & 5 & 2 \\ \hline
A/V equipment & 250 & 50 \\ \hline
Printer & 1000 & 200 \\ \hline
\end{tabular}
\end{center}

Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays, (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day.

Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.

We lack good data about the campus network, save for the existence of approximately 4,000 wifi access points. Based on this, we choose to model network emissions as a configurable percentage of device (local, desktop, server) emissions, by default 10\%.

\subsubsection*{Embodied emissions}

For several device types, especially end-user devices such as phones and laptops, embodied emissions
comprise a majority of lifetime emissions. Different sources report widely different values for these emissions. Our approach has been to find several studies and industry references, and then estimate a mean reference value as well as likely minimum and maximum values.

The following table summarizes results from the literature.

\begin{center}
\begin{tabular}{|l|r|l|}
\hline
& \textbf{Embod. emiss.} & \\
\textbf{Device type} & \textbf{(\kgcoe)} & \textbf{Source} \\ \hline
Laptop & 121 & \textcite{ecoinvent}  \\ \hline
Laptop & 104 & \textcite{teehan2013} (low) \\ \hline
Laptop & 338 & \textcite{teehan2013} (high) \\ \hline
Laptop & 244 & \textcite{rarecoil} \\ \hline
Laptop & 202 & \textcite{unctadder2024} \\ \hline
Desktop & 238 & \textcite{ecoinvent} \\ \hline
Desktop & 303 & \textcite{teehan2013} \\ \hline
Desktop & 403 & \textcite{unctadder2024} \\ \hline
Desktop & 289 & \textcite{dellpcf} \\ \hline
Desktop & 277 & \textcite{boavizta} \\ \hline
Server & 383 & \textcite{teehan2013} \\ \hline
Server & 1252 & \textcite{davy2021} (Dell) \\ \hline
Server & 1912 & \textcite{davy2021} (EC2) \\ \hline
Server & 899 & \textcite{boavizta} (medium server)\\ \hline
Phone & 68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
Phone & 54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
Phone & 50 & \textcite{unctadder2024} \\ \hline
Phone & 50 & \textcite{lovehagen2023} \\ \hline
\end{tabular}
\label{tab:embodied_emissions}
\end{center}

For laptops, the numbers from Ecoinvent, a reputable proprietary Swiss life cycle inventory database used in the UZH sustainability report, appear to be on the low end. The most comprehensive publicly available data comes from \textcite{rarecoil}, which lists manufacturer information for over 90 popular devices. Mean supply chain emissions in this data set are 244 \kgcoe\ with a standard deviation of 128 kg. To be conservative, we split the difference between this number and the Ecoinvent values and model laptops embodied emissions at 180 \kgcoe\ / device.

For desktops, most estimates are in the range of 250 - 400 \kgcoe. The most detailed data comes from Dell's Product Carbon Footprints (\cite{dellpcf}), which the company publishes for each of their products. A sample of 15 desktops has mean supply chain emissions of 289 \kgcoe\ with $\sigma = 80$ kg. We model desktop embodied emissions at 288 \kgcoe\ / device.

In the case of server, \textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances. The former have mean supply chain emissions of 1252 \kgcoe\ with $\sigma = 330$ kg, the latter 1912 kg with $\sigma = 885$ kg. \textcite{boavizta} reports mean supply chain emissions of 899 \kgcoe\ for a medium-sized server. We model server supply chain emissions at 1252 \kgcoe\ / device.

The range of values for mobile phones is much smaller. We model phone embodied emissions at 50 \kgcoe\ / device.

We found less data for other types of devices. For now, we model them as follows:

\begin{itemize}
    \item Computer monitors: 344 \kgcoe\ (\cite{teehan2013}, \cite{dellpcf}).
    \item Conference room displays: 753 \kgcoe\ (scaling monitor by $(40''/27'')^2$). We use this as a proxy for all audiovisual equipment (lecture room projectors, etc.). Note that the numbers are sufficiently low that errors in this estimate do not significantly affect the overall result.
    \item Printer/copier stations: 1167 \kgcoe\ (\cite{ecoinvent}).
    \item Network equipment: We did not find reliable lifecycle assessments for routers and switches,
      and have little information about UZH network architecture. The vast majority ($80-95\%$) of
      lifecycle emissions of network gear stem from usage (\cite{cisco2024}, \cite{jacob2023}).
      For now, we model the embodied emissions of network equipment like we do its operational emissions, namely as a 10\% overhead on the number of devices (laptops, desktops, servers).
\end{itemize}

\subsection{Cloud and supercomputing facilities}

\subsection{LLMs}

We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
to cover the rest of the university population.

For each device instance, we pick embodied emissions, power draw, and duty cycles either
deterministically or by sampling from the appropriate distribution.

Combining power draw and duty cycle estimates with ground-truth data about the Swiss electricity mix,
we estimate annual operational emissions and add them to the embodied emissions scaled by the device
type's expected lifetime.

For on-campus data centers, for which presumably there are electricity bills or other measures of
overall energy use, we would like to use that data to evaluate our model.

A possible future refinement will be to model the duty cycle as a function of time of day
and incorporate finer-grained electricity mix data.

\subsection{Cloud emissions}

We model cloud emissions based on the assumptions and data described in
Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
more accurate data.

\subsection{LLM emissions}

We model LLM emissions by sampling from the per-capita daily query distribution and computing
the carbon intensity of each query based on the average European electricity mix. While there
are many uncertainties about the level of LLM adoption, there appears to be reasonable consensus
in the literature about the energy consumption of each query.

\subsection{Cloud data centers}
\label{sec:cloud_data_centers}

The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
infrastructure. The best resource we have been able to find is the Datavizta API (\cite{boavizta}).

Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as follows:
\begin{itemize}
    \item One-for-one replacement of one campus server with one Azure D8S\_v3 instance (2x Intel
      Xeon, 8 cores, 32 GB RAM, 2TB SSD) (\cite{msftvms}).
    \item Deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region .
    \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 \kgcoe\ for usage,
      97 \kgcoe\ for manufacturing.
\end{itemize}

\subsection{Large language models}

Energy consumption of LLMs is being studied extensively (\cite{budennyy2022eco2ai},
\cite{castano2023exploring},
\cite{devries2023growing},
\cite{gowda2024watt},
\cite{harding2024watts},
\cite{heguerte2023estimate},
\cite{luccioni2022estimating},
\cite{luccioni2023counting},
\cite{patterson2021carbon},
\cite{rodriguez2024evaluating},
\cite{tripp2024measuring}).

The best estimates are that use of large language model (LLM) like ChatGPT averages 3-4 Wh / request,
or approximately 10x the energy of a traditional search engine query. This number appears to be decreasing
rapidly as model hardware and software improve, even though overall energy use is increasing due to
increased utilization. Nevertheless, for the model we assume that queries consume 3 Wh each.
We assume that on average every user will issue 15 queries per day.

LLMs are distributed and the location that serves any particular request is unknown, so we use the
average European electricity mix to compute carbon intensity.

The embodied emissions and training emissions of LLMs are difficult to estimate and attribute precisely.
However, \textcite{devries2023growing} provides the following information about ChatGPT:
\begin{itemize}
    \item Operational power consumption of 500 MWh / day.
    \item Approximately 1300 MWh consumed in training.
    \item Approximately 4000 servers.
\end{itemize}

Assuming a relatively clean grid at 200 \gcoekwh (a low estimate), and that each server has embodied emissions of 4 t \coe\ (a high estimate),
we obtain:
\begin{itemize}
    \item 100 t \coe\ / day operational emissions.
    \item 260 t \coe\ training emissions.
    \item 16 kt \coe\ total embodied emissions.
\end{itemize}

After one year of operation, operational emissions are approximately 36 kt \coe, whereas training and embodied emissions are about 16 kt \coe,
so operational emissions account for $\approx 70\%$ of the total. With less conservative assumptions (dirtier grid, lower embodied emissions),
the proportion of operational emissions increases further.

As a result, when modeling overall LLM emissions, we multiply estimated operational emissions by 1.4 to account for training and embodied emissions.
We acknowledge that this is, at best, a rough estimate based on just one LLM.

\subsection{Electricity carbon intensity}

The Swiss Federal Office for the Environment published an extensive report (\cite{krebs2018umweltbilanz})
on the environmental impact of electricity generation in Switzerland in 2018.

However, for the purposes of the model, we use data from \textcite{electricitymaps}. Specifically,
we use the carbon emissions for electricity {\em consumption} (not generation) in Switzerland for 2023
at hourly granularity, then aggregate the data into two-hour blocks, and compute the mean of each two
hour block over the course of the year. This gives us twelve data points that describe average
carbon intensity variation over the course of the day.

\subsection{Model validation}

Using Ecoinvent embodied emissions values where available, and only the numbers and types of devices reported in the UZH 2024 Sustainability Report~\cite{uzh:sustainability:report}, our model estimates emissions of 1,460 t \coe, within 1\% of the 1,454 estimated in the UZH report, even though the methodology is somewhat different. Our model exposes average device power ratings and duty cycle, grid emissions intensity, etc., directly, whereas the approach used in the sustainability report relies on gross emissions factors (\kgcoe/device for embodied emissions, \kgcoe/device/day for operational emissions) provided by Ecoinvent.

While this is admittedly just one data point, it provides some confidence that our assumptions are realistic and that the model can be used to extrapolate emissions and trends beyond the scope of the sustainability report.

\subsection{Limitations and validation}

The biggest limitation is the modest availability of some ground truth data about UZH IT infrastructure. Specifically, we have only estimates of campus data center energy use, and limited information abut UZH research workloads (whether AI or not). We assume that many research AI workloads will be relatively similar in intensity to existing scientific computation workloads. We also have limited information (e.g., peak power consumption) about UZH use of external high-performance compute resources such as Alps/CSCS.

Additionally, cloud vendors and AI companies are secretive about their infrastructure. We have relied on public datasets and research literature to estimate cloud and AI emissions.

Despite these limitations, our bottom-up model aligns closely with two UZH data points. First, the model correctly estimates approximately $400$ kW of compute for central campus data centers. Second, for university-owned campus IT infrastructure, the model estimates $1.4$ kt \coe\ emissions per year, close to the $1.45$ kt reported in the 2024 university sustainability report \cite{uzh:sustainability:report}.

\section{Results}
\label{sec:results}

Our analysis shows that manufacturing (“embodied”) emissions dominate UZH's IT carbon footprint, while operational impacts, including AI adoption, are relatively modest.

\begin{itemize}

\item Under the assumption that the vast majority (90\%) of UZH students and staff have a laptop and smartphone, manufacturing personal devices accounts for over half of UZH's IT carbon footprint. For the estimated 32,000 laptops and smartphones, embodied emissions total $\sim 2400$ t \coe\ annually. Usage-related emissions amount to about a tenth of that.

\item Widespread adoption of LLMs and other frontier models should have moderate impact on total emissions. If everyone at UZH made heavy use of AI (15 queries daily), this would add approximately 220 t CO2e annually, of which 160 t directly attributable to queries. While this would almost double an individual's real-time energy use, the impact on overall emissions is relatively small.

\item Embodied emissions are an important consideration for university-owned devices as well. If we exclude personal devices and LLMs, then total IT emissions shrink to $\sim 1500$ t \coe, of which $\frac{2}{3}$ comprises the embodied emissions of classroom computers, wall displays, printers, etc.

\item Data center cooling and efficiency represents a potential savings opportunity. The largest operational emissions are from data center servers and cooling (320 t). UZH data centers currently require as much energy for cooling as for computation, compared to the $6-15\%$ cooling overhead of cloud industry leaders. Moving workloads to efficient cloud providers could reduce data center emissions from $\sim 560$ t to $\sim 135$ t \coe\ annually, though this estimate requires further validation.

\end{itemize}

\section{Recommendations}
\label{sec:recommendations}

Do not panic about LLMs. Their emissions comprise a small fraction of total IT emissions, they will get more efficient, and attempting to curb their use is likely to be a losing battle.
Invest in infrastructure for the long term. Run awareness campaigns and adapt IT policies so that personal devices and university IT systems are used for as long as possible. Amortize manufacturing emissions over the longest period possible.
Improve the cooling efficiency of university data centers, or engage cloud vendors and academic compute centers (such as CSCS) to validate the potential benefits of a migration to cloud.



\section{Related work}
\label{sec:related:work}



\section{Conclusion}

\printbibliography

\end{document}
